---
title: "Hard-Filtered v. Imputed Variants"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 4
editor_options:
  chunk_output_type: console
---

---
```{r setup, message=FALSE, warning=FALSE, include=FALSE}
require(tidyverse)
require(tidymodels)
require(nationalparkcolors)
require(RColorBrewer)
require(GenomicRanges)
```

Analysis date: `r format(Sys.time(), '%B %d, %Y')`

```{r message=FALSE, warning=FALSE, include=FALSE}
setwd("~/Documents/projects/NemaScan_Performance/")
load(file = "data/NemaScan_Performance.CeNDR2020_PowerPrecision_GAMMA_EMMA.20200924.RData") # EMMA, hard filtered variants
hard.filtered <- simulation.metrics.df %>%
  tidyr::separate(col = sim,
                  into = c("nQTL","Rep","h2","MAF","effect_range","strain_set"), 
                  sep = "_", remove = F) %>%
  tidyr::separate(col = QTL,
                  into = c("CHROM","POS"), 
                  sep = ":", remove = F) %>%
  dplyr::mutate(h2 = as.factor(h2),
                POS = as.numeric(POS),
                startPOS = as.numeric(startPOS),
                peakPOS = as.numeric(peakPOS),
                endPOS = as.numeric(endPOS),
                var.exp  = as.numeric(var.exp),
                peak_id = as.numeric(peak_id),
                BETA = as.numeric(BETA),
                Effect = as.numeric(Effect),
                Frequency = as.numeric(Frequency),
                log10p = as.numeric(log10p),
                interval_size = as.numeric(interval_size)) %>%
  dplyr::filter(algorithm != "No Successful Mappings") %>%
  dplyr::mutate(variant.set = "hard-filtered")

load(file = "data/NemaScan_Performance.CeNDR2020_PowerPrecision_GAMMA.20200909.RData") # EMMAx, imputed variants
imputed <- simuation.metrics.df %>%
  tidyr::separate(col = sim,
                  into = c("nQTL","Rep","h2","MAF","effect_range","strain_set"), 
                  sep = "_", remove = F) %>%
  tidyr::separate(col = QTL,
                  into = c("CHROM","POS"), 
                  sep = ":", remove = F) %>%
  dplyr::mutate(h2 = as.factor(h2),
                POS = as.numeric(POS),
                startPOS = as.numeric(startPOS),
                peakPOS = as.numeric(peakPOS),
                endPOS = as.numeric(endPOS),
                var.exp  = as.numeric(var.exp),
                peak_id = as.numeric(peak_id),
                BETA = as.numeric(BETA),
                Effect = as.numeric(Effect),
                Frequency = as.numeric(Frequency),
                log10p = as.numeric(log10p),
                interval_size = as.numeric(interval_size)) %>%
  dplyr::filter(algorithm != "No Successful Mappings") %>%
  dplyr::mutate(variant.set = "imputed")

dat <- hard.filtered %>%
  dplyr::full_join(., imputed) %>%
  dplyr::filter(algorithm == "LMM-EXACT") %>%
  droplevels()
```


## Simulation Parameters

NemaScan simulation performance was assessed with the following experimental parameters:

* ###### Number of Simulated QTL: ```r levels(as.factor(dat$nQTL))```
* ###### Sample Population(s): ```r levels(as.factor(dat$strain_set))```
* ###### Heritability(ies): ```r levels(as.factor(dat$h2))```
* ###### MAF(s): ```r levels(as.factor(dat$MAF))```
* ###### Number of Replicates per Regime: ```r max(as.numeric(dat$Rep))```
* ###### QTL Effect Range: ```r levels(as.factor(dat$effect_range))```
* ###### Algorithm: ```r levels(as.factor(dat$algorithm))```
* ###### Variant Set: ```r levels(as.factor(dat$variant.set))```

## Simulated QTL Locations
```{r simulated QTL locations, echo=FALSE}
dat %>%
  dplyr::filter(Simulated == TRUE) %>%
  dplyr::filter(!is.na(CHROM)) %>%
  ggplot(mapping = aes(x = POS/1000000, y = Effect)) + 
  theme_bw() + 
  geom_point() + 
  facet_grid(variant.set~CHROM, drop = TRUE) + 
  labs(x = "Genomic position (Mb)",
       y = "Causal QTL Effect",
       title = "Where were QTL Simulated?")

dat %>%
  dplyr::filter(Simulated == TRUE) %>%
  dplyr::filter(!is.na(CHROM)) %>%
  ggplot(mapping = aes(x = Frequency, fill = variant.set)) +
  theme_bw() + 
  geom_density(alpha = 0.6) + 
  scale_fill_manual(values = nationalparkcolors::park_palettes$RockyMountains[c(2,4)], name = "Variant Set") + 
  theme(legend.position = "top") + 
  xlim(c(0.05,0.5)) + 
  labs(x = "Minor Allele Frequency",
       y = "Frequency of Simulated QTL (Smoothed Density)")

dat %>%
  dplyr::filter(Simulated == TRUE) %>%
  dplyr::filter(!is.na(CHROM)) %>%
  ggplot(., mapping = aes(x = abs(Effect), fill = variant.set)) +
  theme_bw() + 
  geom_density(alpha = 0.6) + 
  scale_fill_manual(values = nationalparkcolors::park_palettes$RockyMountains[c(2,4)], name = "Variant Set") + 
  theme(legend.position = "top") + 
  labs(x = "abs(QTL Effect)",
       y = "Frequency of Simulated QTL (Smoothed Density)")
```

## Performance Analysis

```{r power precision resolution, echo=FALSE}
power <- dat %>%
  dplyr::group_by(h2, variant.set) %>%
  yardstick::sens(truth = Simulated, estimate = Detected) %>%
  dplyr::mutate(.metric = "Power")

precision <- dat %>%
  dplyr::group_by(h2, variant.set) %>%
  yardstick::precision(truth = Simulated, estimate = Detected) %>%
  dplyr::mutate(.metric = "Precision")

resolution <- dat %>%
  dplyr::filter(Detected == TRUE) %>%
  dplyr::filter(!is.na(var.exp)) %>%
  dplyr::select(CHROM, POS, Simulated, Frequency, interval_size, algorithm, h2, var.exp, variant.set) %>%
  dplyr::group_by(h2, variant.set) %>%
  dplyr::summarise(mean(interval_size), median(interval_size), sd(interval_size), 
                   mean(var.exp), median(var.exp), sd(var.exp))

summarized <- power %>%
  dplyr::full_join(., precision) %>%
  tidyr::pivot_wider(names_from = .metric, values_from = .estimate) %>%
  dplyr::left_join(.,resolution) %>%
  dplyr::mutate(mean.interval.Mb = `mean(interval_size)`/1000000,
                median.interval.Mb = `median(interval_size)`/1000000,
                sd.interval.Mb = `sd(interval_size)`/1000000, 
                mean.PVE = `mean(var.exp)`*100,
                median.PVE = `median(var.exp)`*100,
                sd.PVE = `sd(var.exp)`*100, 
                .keep = "unused")
```

```{r ROC curve, echo=FALSE}
# CAVEAT: TNR = 0, CIs are relevant and we don't want to exclude tagged variants
dat %>%
  dplyr::group_by(h2, variant.set) %>%
  yardstick::roc_curve(Simulated, log10p) %>%
  dplyr::mutate(h2 = as.numeric(h2)) %>%
  ggplot(., mapping = aes(x = 1-specificity, y = sensitivity, group = h2, colour = variant.set, alpha = h2)) + 
  theme_bw() + 
  geom_path() +
  geom_abline(lty = 3) +
  scale_colour_manual(values = nationalparkcolors::park_palettes$RockyMountains[c(2,4)], name = "Variant Set") + 
  facet_grid(.~variant.set, scales = "free", space = "free") +
  labs(x = "False-Positive Rate", y = "Power")
```

### Mapping Performance with Imputed Variants vs. Hard-Filtered Variants

```{r power precision plots, echo=FALSE}
summarized %>%
  dplyr::mutate(h2 = as.numeric(levels(h2))[h2]) %>%
  ggplot(., mapping = aes(x = h2, y = Power, fill = variant.set)) +
  theme_bw() +
  geom_jitter(width = 0.01, height = 0, shape = 21, size = 3) +
  geom_line(aes(group = variant.set)) +
  scale_fill_manual(values = nationalparkcolors::park_palettes$RockyMountains[c(2,4)], name = "Variant Set") + 
  ylim(c(0,1)) + 
  labs(title = "Power v. Heritability",
       x = expression(italic(h^2)))

summarized %>%
  dplyr::mutate(h2 = as.numeric(levels(h2))[h2]) %>%
  ggplot(., mapping = aes(x = h2, y = Precision, fill = variant.set)) +
  theme_bw() +
  geom_jitter(width = 0.01, height = 0, shape = 21, size = 3) +
  geom_line(aes(group = variant.set)) +
  scale_fill_manual(values = nationalparkcolors::park_palettes$RockyMountains[c(2,4)], name = "Variant Set") + 
  ylim(c(0,1)) + 
  labs(title = "Precision v. Heritability",
       x = expression(italic(h^2)))
```

### Mapping Resolution with Imputed Variants vs. Hard-Filtered Variants

```{r interval width plots, echo=FALSE}
summarized %>%
  dplyr::mutate(h2 = as.numeric(levels(h2))[h2]) %>%
  ggplot(., mapping = aes(x = h2, y = mean.interval.Mb, fill = variant.set)) +
  theme_bw() +
  geom_pointrange(aes(ymin = mean.interval.Mb-sd.interval.Mb,
                      ymax = mean.interval.Mb+sd.interval.Mb), alpha = 0.7, shape = 21, size = 1) +
  scale_fill_manual(values = nationalparkcolors::park_palettes$RockyMountains[c(2,4)], name = "Variant Set") + 
  labs(title = "QTL Resolution v. Heritability",
       x = expression(italic(h^2)),
       y = "Mean QTL Interval Width (Mb)")

summarized %>%
  dplyr::mutate(h2 = as.numeric(levels(h2))[h2]) %>%
  ggplot(., mapping = aes(x = Power, y = mean.interval.Mb, fill = variant.set)) +
  theme_bw() +
    geom_pointrange(aes(ymin = mean.interval.Mb-sd.interval.Mb,
                        ymax = mean.interval.Mb+sd.interval.Mb), alpha = 0.7, shape = 21, size = 1) +
  scale_fill_manual(values = nationalparkcolors::park_palettes$RockyMountains[c(2,4)], name = "Variant Set") + 
  labs(title = "QTL Resolution v. Mapping Power",
       x = "Power",
       y = "QTL Interval Width (Mb)")
```

### Effect Estimation with Imputed Variants vs. Hard-Filtered Variants

```{r power precision v varexplained, echo=FALSE}
summarized %>%
  dplyr::mutate(h2 = as.numeric(levels(h2))[h2]) %>%
  ggplot(., mapping = aes(x = Power, y = mean.PVE, fill = variant.set)) +
  theme_bw() +
  geom_pointrange(aes(ymin = mean.PVE-sd.PVE,
                      ymax = mean.PVE+sd.PVE), alpha = 0.7, shape = 21, size = 1) + 
  scale_fill_manual(values = nationalparkcolors::park_palettes$RockyMountains[c(2,4)], name = "Variant Set") + 
  labs(title = "Power v. QTL Variance Explained",
       y = "Variance Explained by QTL (%)",
       x = "Mapping Power")

summarized %>%
  dplyr::mutate(h2 = as.numeric(levels(h2))[h2]) %>%
  ggplot(., mapping = aes(x = Precision, y = mean.PVE, fill = variant.set)) +
  theme_bw() +
  geom_pointrange(aes(ymin = mean.PVE-sd.PVE,
                      ymax = mean.PVE+sd.PVE), alpha = 0.7, shape = 21, size = 1) + 
  scale_fill_manual(values = nationalparkcolors::park_palettes$RockyMountains[c(2,4)], name = "Variant Set") + 
  labs(title = "Precision v. QTL Variance Explained",
       y = "Variance Explained by QTL (%)",
       x = "Mapping Precision")
```
